# Understanding our spatial data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

* Explain e.g. cosmx in detail, refer to similarities in e.g. xenium
* Load data into object
* Brief explanation of different data structures 
* Sample annotation: Applying sample names to individual slides. Mention difficulty of multiple samples per slide (can we recommend some approach?)

What data do you get? How do these differ between CosMX and Xenium?

High-resolution molecular data:

- Expression matrix
- Transcript coordinates
- Read quality

Spatial and imaging data:

- Cell segmentations
- Image files

## How the data was generated

## The raw data

For a cosmx SMI slide (like this experiment).

* **Flat files** : Most of what we need to do processing. Typically a directory of flat files is read in as a whole, and we don't need to worry too much. But note there are no images in here.
  * SLIDE-polygons.csv.gz  : Cell borders
  * SLIDE_exprMat_file.csv.gz : Counts of genes per cell (Counts matrix)
  * SLIDE_fov_positions_file.csv.gz : Location of FOVs on slide
  * SLIDE_metadata_file.csv.gz : Cell level QC metadata
  * SLIDE_tx_file.csv.gz : Location of individual transcripts.

* **Raw files** : Giant ugly directory with lots of files including microscopy images.
  * RawFiles/SLIDE/RUN_CODE/CellStatsDir/Morphology2D : Location of images
  * https://github.com/Nanostring-Biostats/CosMxDACustomModules/blob/main/Export/CosMxDAExportSetup.docx  
  

For a Xenium slide there is just the one typical output directory; they discuss the formats in the section on [data archiving](https://www.10xgenomics.com/support/software/xenium-onboard-analysis/latest/analysis/xoa-output-archive-data), but again, software tools take the directory as a whole.

What is a FoV?

## Note on `Seurat::LoadNanostring()`

Load the libraries required for this step.

```{r echo=F, warning=F, results=F, warning=F}
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(here)
```

To load the data, we can use the `Seurat::LoadNanostring()` function. However,
this default method drops most of the metadata in the Seurat object. For example,
what field of view (FoV) each cell a member of is missing. 

We provide an alternate function. In time, this should be fixed within Seurat.
See comments [here](https://github.com/satijalab/seurat/discussions/9261)

```{r eval=F}
source(here("scripts/LoadNanostring_edited_function.R"))
example_path <- here("raw_data", "GSM7473682_HC_a")
example_so <- LoadNanostring(example_path, assay='RNA', fov="GSM7473682_HC_a")
```

```{r eval=F}
example_so
```

```console
An object of class Seurat 
999 features across 42045 samples within 1 assay 
Active assay: RNA (999 features, 0 variable features)
 1 layer present: counts
 1 spatial field of view present: GSM7473682_HC_a
```

Explain what's missing

```{r}
so$condition       <- "Healthy Controls"
so$tissue_sample   <- "HC_A" # denotes healthy control, from sample A
```

## Loading all samples from raw data

We use a subsampled data for this workshop, for three healthy control samples (HC)
and three Crohn's disease samples (CD).

The first four field of views are used.

```{r, eval=F}
# How was this created?
seurat_file_00_raw <- here("data", "GSE234713_CosMx_IBD_seurat_00_raw.RDS")
so_raw_full <- readRDS(seurat_file_00_raw)
so_raw_subset <- so_raw_full[, so_raw_full$group %in% c("CD", "HC") & so_raw_full@meta.data$fov <= 4]
saveRDS(so_raw_subset, here("data", "GSE234713_CosMx_IBD_seurat_00_raw_subsampled.RDS"))
```

## Loading the workshop data

TODO: Add image for all FOVs, and outline the ones we are using

```{r}
so_path <- here("data", "GSE234713_CosMx_IBD_seurat_00_raw_subsampled.RDS")
so <- readRDS(so_path)
```

### Understanding the `SeuratObject`

:::{.callout-info}
#### Why Seurat?

Seurat objects are a data structure to store the count data and additional metadata.
As the data is very high dimensional, it is important
that connected information about cells, samples, and data stay connected
during data analysis.

Other frameworks such as `scverse`, and data classes like `anndata` and 
`SpatialFeatureExperiment` represent the data differently, but all meet the
needs above. Differences in compatibility with other tools.

Flexibility across single-cell transcriptomics.
:::

Display the Seurat object.

```{r eval=F}
so
```

**999 features across 68624 samples with 2 assays**

- 980 genes + 19 negprobes
- 68k number of segmented cells analysed
- two different types of measurements, detailed below

**Active assay: RNA (980 features, 0 variable features):**

- RNA assay: Primary gene expression measurement data (counts)
- 980 transcripts measured
- no variable features as feature selection has not been conducted yet

**6 layers present: counts.GSM7473682_HC_a, counts.GSM7473683_HC_b, 
counts.GSM7473684_HC_c, counts.GSM7473688_CD_a, counts.GSM7473689_CD_b, 
counts.GSM7473690_CD_c**

- Each layer represents a sample from either HC or CD donors. Required for
comparison of both groups downstream

**1 other assay present: negprobes**

This secondary assay specifically stores the counts corresponding to negative
control probes. In spatial transcriptomics, tracking these control probes is
crucial for quality control, specificity assessment, and defining background
signal.

**6 spatial fields of view present: GSM7473682_HC_a GSM7473683_HC_b GSM7473684_HC_c GSM7473688_CD_a GSM7473689_CD_b GSM7473690_CD_c:**

The list of six spatial fields of view (FOVs) explicitly links the spatial/image
information stored in the Seurat object to the six specific samples/layers
identified above. This demonstrates that the data set is ready for spatially
resolved analysis and visualization, allowing results (like cell clusters or
gene expression) to be mapped back onto the original tissue sections for all
six integrated samples.
    

### Understanding the Seurat metadata

Display the metadata

```{r}
head(so@meta.data)
```

| Column         | Definition                                 |
| -------------- | ------------------------------------------ |
| Row name       | The sample and cell                        |
| `orig.ident`   | Sample                                     |
| `nCount RNA`   | The number of sequenced transcripts        |
| `nFeature_RNA` | The number of transcripts mapped to a gene |

Each sample should be annotated with its experimental details. This is important
so we know which groups to compare downstream, i.e. case vs. control comparison.
This particular study has one sample per slide (easy!), but there are typically
more. 

Explain input data, what it represents, how it will be used for analysis.
Define terms like centroids, segments, fov

Load a subsampled dataset for working with multiple samples
That takes too long, so loading a preloaded object
Add figure that represents the subset

Explain: Subset data, how many samples and conditions. Represent graphically
e.g. three healthy controls (HC) vs. three CD

Display the metadata columns:

```{r}
names(so@meta.data)
```

There is a lot of information. Use something like skimr for EDA and to
understand each field at a glance

```{r}
skimr::skim(so@meta.data)
```

Explain: number of rows (68624) represent data points (?)
Identify important metadata columns and discuss what they mean

This dataset is a subset of the experimental data - only the first 4 fovs of
each, and only the CD and HC sample groups. 

seurat fov = slide
bruker cosmx fov = region on slide


View the tissue_sample metadata to prepare for the next step

```{r}
table(so$tissue_sample)
```

## Visualising our spatial data

Subset one of the tissue samples `HC_a` for plotting.

```{r}
so_hca <- so[, so$tissue_sample=="HC_a"]
```

Some warnings with not validating FOV, centroid, and seurat objects - are these
important?

```{r}
so_hca
```

Note: layer and FOVs are now only for a single sample, HC_a

Visualise clusters in a spatial context

Explain: key arguments

By segments

```{r}
ImageDimPlot(so_hca,
             fov = "GSM7473682_HC_a",
             axes = TRUE,
             border.color = "white", 
             border.size = 0.1,
             group.by = "fov_name",
             boundaries = "segmentation",
             crop = TRUE,
             nmols = 10000)
```

By centroids

```{r}
ImageDimPlot(so_hca,
             fov = "GSM7473682_HC_a",
             axes = TRUE,
             border.color = "white", 
             border.size = 0.1,
             group.by = "fov_name",
             boundaries = "centroids",
             crop = TRUE,
             nmols = 10000)
```

By an individual gene.

Do not need fov_name this time

```{r}
#And and individual gene
ImageDimPlot(so_hca,
             fov = "GSM7473682_HC_a",
             axes = TRUE,
             border.color = "white",
             border.size = 0.1,
             #group.by = "fov_name",
             boundaries   = "segmentation",
             molecules = "S100A6",
             crop = TRUE,
             nmols = 10000)
```

## Count data

Where are my counts?

```{r eval=F}
#GetAssayData(so, assay = "RNA", layer = "counts")
```

Oh no an error:

``` 
#Error in `GetAssayData()` at SeuratObject/R/seurat.R:1901:3:
#  ! GetAssayData doesn't work for multiple layers in v5 assay.
```

Seurat *5* vs Seurat 4. Multiple assays fovs.
SOme operations only work on a single
Join layers / split layers.
tHis is useful for later steps, but annoying now.

```{r}
#so <- JoinLayers(so)
```

```{r}
#so@assays
```


```{r eval=F}
#GetAssayData(so, assay = "RNA", layer = "counts")
```
