# Quality control and filtering

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries for this step
library(Seurat)
library(tidyverse)
library(here)
```

## Overview

* Basic QC plots + filtering (20min)
* (Negative probes if using cosmx)
* Identify low quality cells

## QC metrics

Read in subsampled seurat object

```{r}
so_path  <- here("data", "GSE234713_CosMx_IBD_seurat_00_raw_subsampled.RDS")
so <- readRDS(so_path)
```

### Identifying potential technical artefacts

Library size (total transcripts) and cell areas are good indicators of technical
artefacts introduced during data generation steps. 

Possible artefacts:
- low quality cells
- cell segmentation challenges (combining two cells, not aligned with real cell
boundaries, leading to high area and low RNA count)

Understanding these can help with biological interpretation downstream,
for example, discerning whether differences in cells are due to e.g. cell type
differences, or technical artefacts.

```{r warning=F}
# Total counts per cell
ggplot(so@meta.data, aes(x=nCount_RNA, col=orig.ident)) +
  geom_density() +
  scale_x_log10() +
  theme_bw() +
  annotation_logticks(sides = "b") +
  ggtitle("Transcript counts per cell")
```

The transcripts per cell all display a log-normal distribution with shifting medians.

Indicates no largely low-quality cells, and a good example of library size effects that should be normalised downstream, prior to analyses.

```{r warning=F}
# Total counts per cell
ggplot(so@meta.data, aes(x=Area, col=orig.ident)) +
  geom_density() +
  scale_x_log10() +
  theme_bw() +
  annotation_logticks(sides = "b") +
  ggtitle("Area (μm^2) per cell")
```

Areas also follow log-normal distributions with medians ranging between 3,000 -
4,000 $\mu m^2$. Sample/FOV `CD.c` has slightly smaller cell area with a median
of ~2,000 $\mu m^2$.

Useful to view bivariate metrics - the larger the cell, the more transcripts, in
general.

```{r, warning=F}
ggplot(so@meta.data, aes(x=Area, y=nCount_RNA)) +
  geom_point(size = 0.1, alpha = 0.4) +
  scale_x_log10() +
  scale_y_log10() +
  annotation_logticks(sides = "bl") +
  facet_wrap(~orig.ident) +
  theme_light()
```


**Exercise:** Display density of negprobes

**Tip:** View the metadata columns using `names(so@meta.data)`, and locate the
name that corresponds to the number of negative probes.

**Tip:** Use the previous code block and replace `nCount_RNA` with
`nCount_negprobes`. Remember to update the `ggtitle()` label.

```{r warning=F}
ggplot(so@meta.data, aes(x=nCount_negprobes, col=orig.ident)) +
  geom_density() +
  theme_bw() +
  ggtitle("Negative probe counts per cell")
```

Suggest to view QC metrics as co-variates - may lack context if displayed alone:

```{r}
so@meta.data %>%
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, colour = tissue_sample)) +
  geom_point(alpha = 0.2, size = 0.8) +
  facet_wrap(~tissue_sample) +
  theme_light() +
  theme(legend.position = "none")
```

The more transcripts detected `nCount_RNA`, the more mapped genes there are. Bad quality cells

Point out where bad quality cells would reside

## FoV QC (CosMX only)

Good to check library sizes across FOVs - red flags include FOVs with close to
no cells.

```{r}
so@meta.data %>%
  unite(col = unique_fov_names, orig.ident, fov, remove = F) %>%
  ggplot(aes(x = nCount_RNA, col = orig.ident, group = unique_fov_names)) +
  geom_density() +
  scale_x_log10() +
  theme_light() +
  annotation_logticks(sides = "b")
```

## Spatial

```{r}
unique(so$orig.ident)
```

```{r}
ImageFeaturePlot(so, fov = "GSM7473684_HC_c", features = "nCount_RNA")
```

```{r}
ImageFeaturePlot(so, fov = "GSM7473684_HC_c", features = "nCount_negprobes")
```

## Filter negative counts per cell

Key metrics:

- `so$percent_neg`: The percentage of negative probes per cell
- `so$mean_neg`: The average negative probes per cell

```{r}
so$percent_neg <- (so$nCount_negprobes / (so$nCount_RNA + so$nCount_negprobes)) * 100

# Join layers to calculate average_neg
so@assays$negprobes <- JoinLayers(so@assays$negprobes)
so$mean_neg <- colMeans(so@assays$negprobes) # Only defined first sample
```

```{r}
# Discuss difference with Xenium - xenium lower counts, lower background.
max_pc_neg         <- 5

ggplot(so@meta.data, aes(y=mean_neg, x=nCount_RNA)) +
  geom_point(alpha=0.1) +
  scale_x_log10() +
  theme_bw() +
  ggtitle("Negative probes vs counts")

min_count_per_cell <- 1
ggplot(so@meta.data, aes(x = nCount_RNA, y = mean_neg)) +
  geom_hline(yintercept = max_pc_neg, lty=3, colour = "red") +
  geom_vline(xintercept = min_count_per_cell, lty=3, colour = "red") +
  geom_point(pch=3, alpha=0.1) +
  scale_x_log10() +
  theme_bw() +
  ggtitle("Negative probes vs counts")
```

```{r}
hc_idents <- unique(so$orig.ident)[1:3]
cd_idents <- unique(so$orig.ident)[1:3]
```

```{r}
ImageFeaturePlot(so, fov = hc_idents, features = c("percent_neg"))
```

```{r}
ImageFeaturePlot(so, fov = cd_idents, features = c("percent_neg"))
```

Note: You can also conduct QC in low dimensional steps following the
normalisation and dimension reduction steps in this workshop. QC is often
iterative. First pass should be permissive with filtering.

For the workshop we will proceed directly with the pre-processing steps.
